A Survey of Scientific Workflow Management Systems

### 2014-04-16 paper.tex
edits

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm that makes them a limited contenders. 
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{Bulk Work.}
Workflows are often used to perform computational work in bulk. This could be
bulk data processing or running multiple distinct processing on a single set of
data. In addition to carrying out such work, workflow environments provide
oppurtunities for giving structure and order to a set of tasks and often with
sufficient expressivity, choices and selections to perform tasks selectively.

\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include composing and marshalling of execution,
browsing and selecting data, visualizing results. Furthermore,
dissemination of methods and results to communities and collaborators are
some of the activities that SWEs make convenient for end-users.

\item \textbf{Resource Access and Utilization.} Computational platforms and
 access modalities change rapidly with upgradation of existing technologies and
adaptation of new techniques. Clouds are the latest examples of new techniques
of computational platform access while GPGPUs examples in upgradation of
existing processing environments. Additionally, with the advent of fast and
dedicated networks such as ESNet and federated visualization resources, remote
visualization is gaining prominence among scientific users. SWEs play important
role as platforms to give transparent and hassle free access to such novel and
multiple environments to users by hiding the technicalities of access methods.

\item \textbf{Repeatable Execution.} Scientific users often require a
repeatable execution of same workflow for reproduction of results and
demonstration of methods. SWEs can record and save workflows, associated
parameters, source data and historical results to aid in such activities.
Additionally, SWEs facilitate an incremental development of such methods
leading to augmentation of past efforts to current and future developments.

\item \textbf{Provenance and MetaData.} Historical data and metadata has been
crucial entities in many research fields. They help connecting results from
a multitude of research aspects. A provenance trace of computation can help
checkpoint methods and reproduce and modify partial computation. Automated
observations of deviations in scientific phenomena from provenance data is one
of the key triggers for new scientific discoveries. Fields of studies involving
genomic and protiens are examples of domains where provenance and metadata
plays key roles. SWEs can be integrated or extended with metadata and
provenance techniques rendering powerful computing traceability features.

\item \textbf{Sharing and Collaboration.} Workflow managers are often targetted
towards a community of users. For instance, the Galaxy workflow system is
targetted specifically for the bioinformatics users. Galaxy attracts many users
from the same community thus promoting sharing and collaboration via a common
computational platform. Another such example is Taverna which is geared towards
sharing and collaborative workflow oriented scientific computation enabling
users to discover new services introduced to a community.

@@

% How SWEs differ from traditional programming languages and runtimes

As a consequence to these design goals, the SWEs differ in many aspects with
respect to traditional programming environments. The following list identifies
the major differences between traditional and the SWE programming
paradigm:

@@

\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
extensive usage of distributed resources than conventional programs. Owing
to the large-scale experiments, for which scientific users turn to
automation, their demands for computational power often exceeds that of a
single personal computer. Such applications are often composed of multiple,
distinct computational stages making them well-suited to the workflow
paradigm. These properties also make them ideal candidates for a
large-scale distributed computing environment. The case for suitability of
SWEs on large-scale distributed infrastructures has been well-argued upon
in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
ensure scalable and optimized performance is vital.
%

\item \textbf{Data Management.} The scale of data transfer is much higher in
SWEs from one execution site to others as well as among storage resources
while it is lower (ranging from variables to small data structures) in
programs.
%

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing
    %Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
experts to utilize the computational power. This makes it essential for
them to have a friendly, and intuitive user-interface for novice and
non-expert users. For this reason, SWE user interfaces are dominated by the
visual paradigm of programming. On the other hand, traditional programs are
most often written by expert programmers. Thus, they have a text-based
composition environment albeit with visually assisting programming
environments such as IDEs. However, they are seldom to the extent that a
non-expert user can graphically compose programs.

@@

We concede that the above is not an exhaustive list of differences in features.
However, it is reasonably inclusive and we believe it covers the breadth of
aspects we cover in the current work. We do not include other interesting SWE
features such as provenance and interoperability in this analysis as they are
out of scope of the objectives of the current paper.

@@

\subsection{Workflow}
Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
Graph (DAG). For the purpose of this text, we define a workflow as a
computation with at least two discrete tasks either forming distinct
computational stages or running for two or more distinct datasets as
parameters. Consequently, a one-off task or running a task for a single copy of
data is \emph{not} a workflow. For the rest of this paper, we will use the above
definition of a workflow.
%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%
\subsection{Architecture of a Generic SWE}
Figure~\ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
components and their interaction in a SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages. The \textit{workflow
language} becomes an intermediate representation for communication between user
interface and the \textit{workflow engine}. The workflow engine acts as a
compiler/interpreter and scheduler for the workflow tasks and interfaces with
the underlying \textit{distributed resources}. It also monitors the workflow
progress and collects the intermediate and final results.

@@

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs, studying many existing systems for distributed computing. This taxonomy
is based, to a large extent on the capabilities of the workflow enactors. It
provides only a limited insight to the SWE aspects considered in the present
paper, especially the workflow languages expressiveness and SWEs interactions
with the distributed computing infrastructures. A more recent work by Goderis
and co-authors \cite{goderis-brooks-etal:2009} studies a topic more closely
related to the workflow language expressiveness: it compares the models of
computations of various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. The recent survey from Stratan et. al.
\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
account the latencies involved during the enactment mechanisms. A small survey
and analysis of file-access characteristics of Data-intensive workflows has
been done by Shibata et. al.~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in \cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,delaitre-kiss-etal:2005,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR DAGMan
\cite{dagman} to abstractions for parallel computations such as Petri nets
\cite{alt-hoheisel:2005}, meta-models \cite{nemo-glatard-etal:2007},
data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
languages such as SwiftScript \cite{swift}. Glatard's PhD thesis
\cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
functional, graph, service-oriented and executable representations. A detailed
classification and taxonomy of parallelism patterns in Grid Workflows has been
presented by Pautasso and Alonso in \cite{pautasso1}. This classification is
more generic while the one presented in Glatard's thesis has emphasis on
service based workflows.  Each of these approaches can be defended through some
aspects well covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of applications logic for nonexpert users and scripting is
extensively used by programmers for prototyping, etc.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG\footnote{http://graal.ens-lyon.fr/diet/workflow.html}
\cite{caron-desprez:2005}, engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. In spite of the presence of many
different approaches to workflow languages, new application requirements
continue to pose workflow expressiveness challenges. The community is yet to
realize a truly generic workflow language that can express a wide variety of
flow patterns encountered in scientific domains.

>>>>

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm which makes them a strong contenders for the
purpose of high-level application programmers.
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{Bulk Work.}
Workflows are often used to perform computational work in bulk. This could be
bulk data processing or running multiple distinct processing tasks on a single
set of data. In addition to carrying out such work, workflow environments
provide oppurtunities for giving structure and order to a set of tasks and
often with sufficient expressivity, choices and selections to carry out a bulk
of work selectively. Some of the examples of such work are \emph{parameter
sweeps} where tasks are run over a series of parameter values, \emph{ensemble
studies} where tasks are carried out over discrete groups of parameters over a
space.

\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include visually formulating a task, composing and
marshalling of workflow, browsing and selecting data, visualizing and sharing
results.

\item \textbf{Resource Access and Utilization.} Computational platforms and
access modalities change rapidly with upgradation of existing technologies and
adaptation of new techniques. Clouds are the latest examples of new techniques
of computational platform access while GPGPUs are examples of upgradation of
existing processing environments. Additionally, with the advent of fast and
dedicated networks such as ESNet and federated visualization resources, remote
visualization is gaining prominence among scientific users. SWEs play
important role as platforms to give transparent and hassle free access to such
novel and multiple environments to users by hiding the technicalities of
access methods and usage and configuration related nuances.

\item \textbf{Repeatable Execution.} Scientific users often require a
 repeatable execution of same application(with same parameters and input data)
 for reproduction of results and demonstration of methods. SWEs can record and
 save workflows, associated parameters, input data and historical results to
 aid in such activities.  Additionally, SWEs facilitate an incremental
 development of these methods leading to augmentation of past efforts to current
 and future developments.

\item \textbf{Provenance and MetaData.} Historical data and metadata has been
crucial entities in many research fields. They help connecting results from
a multitude of research aspects. A provenance trace of computation can help
checkpoint methods and reproduce and modify partial computation. Automated
observations of deviations in scientific phenomena from provenance data is one
of the key triggers for new scientific discoveries. Fields of studies involving
genomic and protiens are examples of domains where provenance and metadata
plays key roles. SWEs can be integrated or extended with metadata and
provenance techniques rendering powerful computing and traceability features.

\item \textbf{Sharing and Collaboration.} Workflow systems are often targetted
towards a particular community of users. Furthermore, dissemination of methods and results
to communities and collaborators are some of the activities that SWEs make
convenient for end-users. For instance, the Galaxy workflow system is targetted
specifically for the bioinformatics users encouraging them to collaborate over
problems and share results. Another such example is Taverna which is geared
towards sharing and collaborative workflow oriented scientific computation
enabling users to discover new services introduced to a community. Sharing and
collaboration of applications and related data have become ubiquitious with the
rise of social web. Initiatives such as
\emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
\emph{gitHub}~\cite{github-web} have enabled easy sharing and collaboration of
reproducible workflows on the social web.

@@

The above described goals also form a basis for the need for a SWE. The need
for the use of SWE can arise as a result of one or more of such requirements.
Traditional programming platforms are not suitable to address these
requirements as they operate at a different level of abstraction. SWEs differ
in many respects among themselves as well as with respect to traditional
programming environments. The following list identifies the major differences
between traditional and the SWE programming paradigm:

@@

\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
extensive usage of distributed resources than conventional programs. Owing
to the large-scale experiments, for which scientific users turn to automation,
their demands for computational power often exceeds that of a single personal
computer. These properties make them ideal candidates for a large-scale
distributed computing environment. The case for suitability of SWEs on
large-scale distributed infrastructures has been well-argued upon
in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
ensure scalable and optimized performance is vital. Additionally, most SWEs are
capable of executing workflows at small scale with minimal tuning. This
multi-scalability enables quick and easy conceptualization and prototyping of
new workflows.
%

\item \textbf{Data Management.} The scale of data transfer is much higher in
SWEs from one execution site to others as well as among storage resources while
it is lower (ranging from variables to small data structures) in programs.
Owing to such scales, applications often need access to special data movement
and access protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
shipped with capabilities to handle such protocols.
%

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing
    %Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interface to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Modern computing infrastructures such as Clouds, GPUs and
supercomputers with hybrid nodes pose new challenges of interface development.
As SWEs involve running many tasks over an infrastructure it becomes crucial to
interface with these systems such that the tasks are scheduled in an optimal
manner. Many SWEs implement
\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
reserve resources via pseudo-tasks that act as placeholders for real tasks.
Additionally, decision to manage working data storage on disc, memory and cache
are some of the distinctions between a traditional programming and workflow
platforms.

\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of computation is
known as dataflow computing. A workflow thus does not follow a fixed order of
execution. Different execution sequences are possible which are dominated by
dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
of execution is coarse-grained, and often termed as ``processor'', ``task'',
``service'' or ``actor'' in the SWE community. This notion of high-level
granularity drives the depth to which a workflow can be expressed in a given
language form. In the rest of the paper, we refer to a unit of execution of a
workflow system as \emph{processor}. On the other hand, a unit of execution in
most programming languages is a much simpler and finer executable statement
that accomplishes a small computational task. This presents a whole new set of
requirements for expressing the processors and how they should interact with
computational resources and operate upon data.

%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
experts to utilize the computational power. This makes it essential for
them to have a friendly, and intuitive user-interface for novice and
non-expert users. For this reason, SWE user interfaces are dominated by the
visual paradigm of programming. On the other hand, traditional programs are
most often written by expert programmers. Thus, they have a text-based
composition environment albeit with visually assisting programming environments
such as IDEs. However, they are seldom to the extent that a non-expert user can
graphically compose programs. While a graphical interface enhances usability, a
text-based workflow environment can arguably give better control to an
experienced user. Consequently, many workflow systems offer a dual interface of
graphical and text medium to compose workflows.

@@

We concede that the above is not an exhaustive list of differences in features.
However, it is reasonably inclusive and we believe it covers the breadth of
aspects we cover in the current work. We do not include other interesting SWE
features such as application predictability and dynamic restructure of DAGs in
this analysis as they are out of scope of the objectives of the current paper.

@@

\subsection{Workflow}
Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
Graph (DAG). For the purpose of this text, we define a workflow as a
computation with at least two discrete tasks either forming distinct
computational stages or running for two or more distinct datasets as
parameters. Consequently, a one-off task or running a task for a single copy of
data is \emph{not} a workflow. For the rest of this paper, we will use the above
definition of a workflow.
%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%
\subsection{Architecture of a Generic SWE}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in a SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow system} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow system}. The workflow system acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the underlying
\textit{distributed resources}. It also monitors the workflow progress and
collects the intermediate and final results.

@@

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs, studying many existing systems for distributed computing. This taxonomy
is based, to a large extent on the capabilities of the workflow enactors. It
provides only a limited insight to the SWE aspects considered in the present
paper, especially the workflow languages expressiveness and SWEs interactions
with the distributed computing infrastructures. A more recent work by Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely
related to the workflow language expressiveness: it compares the models of
computations of various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. The recent survey from Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
account the latencies involved during the enactment mechanisms. A small survey
and analysis of file-access characteristics for data-intensive workflows has
been done by Shibata~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in \cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR DAGMan
\cite{dagman} to abstractions for parallel computations such as Petri nets
\cite{alt-hoheisel:2005},
data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
languages such as Swift script \cite{swift}. Glatard's PhD thesis
\cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
functional, graph, service-oriented and executable representations. A detailed
classification and taxonomy of parallelism patterns in Grid Workflows has been
presented by Pautasso and Alonso in \cite{pautasso1}. This classification is
more generic while the one presented in Glatard's thesis has emphasis on
service based workflows.  Each of these approaches can be defended through some
aspects well covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of application logic for non-expert users and scripting is
extensively used by programmers for prototyping, etc.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG\footnote{http://graal.ens-lyon.fr/diet/workflow.html}
\cite{caron-desprez:2005}, engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. Despite the presence of many
different approaches to workflow systems, new requirements
continue to pose workflow challenges. The community is yet to
realize a truly generic workflow language that can express a wide variety of
flow patterns encountered in scientific domains.
